---
title: "Fungal secondary metabolism determines aphid vector competence"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(deSolve)
library(fields)
library(sfsmisc)
library(extrafont)
library(RColorBrewer)
loadfonts(device="all")
myFont <- "sans" 
```

This R markdown file contains R code used in the epidemiological modelling component of Simon et al. (2026) "Fungal secondary metabolism determines aphid vector competence".

All figures in the paper can be produced using this file.

```{r echo=F,odeRHS}
fullModel <- function(t, y, par)
{
  # extract state variables
  S <- y[1]
  I <- y[2]
  X <- y[3]
  Z <- y[4]
  
  # extract parameters
  mu <- par$mu
  K <- par$K
  r <- par$r
  Gamma <- par$Gamma
  alpha <- par$alpha
  v <- par$v
  a <- par$a
  beta <- par$beta
  eta <- par$eta
  c <- par$c
  M <- par$M
  # new parameter for Z to avoid I
  delta <- par$delta
  
  dS <- mu * (K - (S + I)) - r * I * S / K - (beta / Gamma) * (delta*S/(delta*S + I)) * Z
  dI <- r * I * S / K + (beta / Gamma) * (delta*S/(delta*S + I)) * Z
  
  dX <- a * (X + Z) * (1 - (X+Z)/M) - (alpha/Gamma) * ((v * I)/(S + v * I)) * X + eta * Z - c * X
  dZ <- (alpha/Gamma) * ((v * I)/(S + v * I)) * X - eta * Z - c * Z

  # return
  return(list(c(dS, 
                dI,
                dX,
                dZ)))
}

#
# if type = 1 then return both infection pathways
#    type = 2 then just aerial
#    type = 3 then just aphids
#
rateChangeInfFromODESoln <- function(odeSoln, par, type)
{
  S <- odeSoln['S']
  I <- odeSoln['I']
  Z <- odeSoln['Z']
  
  r <- par$r
  K <- par$K
  beta <- par$beta
  Gamma <- par$Gamma
  delta <- par$delta
  v <- par$v

  if(type==1)
  {
    thisRet <- r * I * S / K + (beta / Gamma) * (delta * S/(delta * S + I)) * Z  
  }
  if(type==2)
  {
    thisRet <- r * I * S / K 
  }
  if(type==3)
  {
    thisRet <- (beta / Gamma) * (delta * S/(delta * S + I)) * Z  
  }
  
  return(thisRet[,1])
}

severityFromLambda <- function(t, thisL, par)
{
  zetaZero <- par$zetaZero
  omega <- par$omega
  mu <- par$mu
  rho <- par$rho
  K <- par$K
  
  nT <- length(t)
  retVal <- numeric(nT)
  if(nT > 1)
  {
    deltaT <- t[2] - t[1]
    for(i in 2:nT)
    {
      thisVal <- 0
      for(j in 1:(i-1))
      {
        thisTau <- t[j]
        thisSev <- 1/(1 + (1/zetaZero - 1) * exp(-omega * thisTau))
        thisRate <- thisL[i - j] / K
        thisVal <- thisVal + thisRate * thisSev * deltaT
      }
      retVal[i] <- thisVal
    }
  }
  return(retVal)
}
```

## Parameterisation of epidemiological model

Report the default parameterisation
```{r defParams}
# parameters
defParamNIV <- list(K = 300,     # carrying capacity for spikes per unit area
                    mu = 0.1,    # rate of production of spikes
                    r = 0.15,    # aerial infection rate
                    Gamma = 1,   # time an aphid feeds for
                    alpha = 0.5, # probability aphid picks up fungus per visit
                    beta = 0.5,  # probability aphid transmits fungus per visit
                    v = 1.2,     # vector preference (fungus-free vectors to infected plants)
                    a = 0.5,     # aphid birth rate  
                    eta = 0.2,   # rate of loss of infectivity 
                    c = 0.2,     # death rate of aphids
                    M = 800,     # "carrying capacity" of aphids per unit area 
                    delta = 1.5, # vector preference (fungus-carrying vectors to healthy plants)
                    # Severity parameters calculated by model fitting below
                    omega = 0.25,      # rate at which severity increases on a spike
                    zetaZero = 0.034)  # initial proportion of spikelets infected

defParamDON <- defParamNIV
# Three changes for DON relative to NIV:
#     - 1% as likely to pick up fungus from infected head as NIV
#     - 50% as likely to choose infected head as NIV 
#     - 50% more transmissible via aerial spread
defParamDON$alpha <- 0.01 * defParamNIV$alpha
defParamDON$v <- 0.5 * defParamNIV$v
defParamDON$r <- 1.5 * defParamNIV$r
defParamDON$delta <- 1.0
```

## Figures 

The following code generates the following figures (in order)

- Extended Data Figure 9. Panel (a)
- Extended Data Figure 9. Panel (b)
- Figure 4. Panel (f)
- Figure 4. Panel (g)
- Extended Data Figure 9. Panel (c)
- Extended Data Figure 9. Panel (d)
- Extended Data Figure 9. Panel (e)
- Extended Data Figure 9. Panel (f)



```{R echo=F,compareNIVDONAlpha,fig.height = 4,fig.align = 'center'}
# initial conditions
yZero <- c(S = 5, 
           I = 1,
           X = 0.1, 
           Z = 0) 

tMax <- 60
tMaxSev <- tMax
for(i in 1:2)
{
  if(i == 1)
  {
    thisP <- defParamNIV
    sType <- "NIV"
  }else{
    thisP <- defParamDON
    sType <- "DON"
  }
  
  odeRet <- data.frame(ode(y = yZero, 
                           times = seq(0,tMax,length.out=101), 
                           func = fullModel, 
                           parms = thisP, 
                           method = "ode23"))
  
  if(i==1)
  {
    odeRetNIV <- odeRet
  }else{
    odeRetDON <- odeRet
  }
  
  if(T)
  {
    tSev <- seq(0,tMaxSev,by=0.1)
    
    thisOmega <- 1/(1 + (1/thisP$zetaZero - 1) * exp(-thisP$omega * tSev))
    thisLambda <- rateChangeInfFromODESoln(odeRet,thisP,1)

    thisLambdaAerial <- rateChangeInfFromODESoln(odeRet,thisP,2)
    thisLambdaAphid <- rateChangeInfFromODESoln(odeRet,thisP,3)

    thisSev <- severityFromLambda(odeRet$time,thisLambda,thisP)

    if(i==1)
    {
      sevNIV <- thisSev
    }else{
      sevDON <- thisSev
    }
  }  
}



par(las=1)
par(mfrow = c(1,2))
par(family=myFont)

plot(odeRetNIV$time,
     odeRetNIV$I,
     ty="l",
     col="darkred",
     lwd=2,
     xlab="t, time (days)",
     ylab="Heads per square metre",
     main="Infected head density",
     ylim=c(0,thisP$K),
     cex.lab=(11/12),
     cex.axis=(10/12),
     cex.main=(10/12))
lines(odeRetDON$time,
      odeRetDON$I,
      lwd=2,
      col="firebrick2")
legend("topleft",
       c("NIV","DON"),
       col=c("darkred","firebrick2"),
       lwd=2,
       lty=1,
       cex=(9/12))

plot(odeRetNIV$time,
     sevNIV,
     ty="l",
     col="darkred",
     lwd=2,
     xlab="t, time (days)",
     ylab="Proportion of infected heads",
     main="Severity on heads",
     ylim=c(0,1),
     cex.lab=(11/12),
     cex.axis=(10/12),
     cex.main=(10/12))

lines(odeRetDON$time,
      sevDON,
      lwd=2,
      col="firebrick2")
if(F)
{
  legend("topleft",
         c("NIV","DON"),
         col=c("darkred","firebrick2"),
         lwd=2,
         lty=1)
}

for(i in 1:2)
{
	if(i == 1)
	{
		odeRet <- odeRetDON
		sType <- "DON"
	}else{
		odeRet <- odeRetNIV
		sType <- "NIV"
	}
    tSev <- seq(0,tMaxSev,by=0.1)
    
    sTmp <- sprintf("Infection rates (%s)",sType)

    thisLambda <- rateChangeInfFromODESoln(odeRet,thisP,1)
    plot(odeRet$time,
         thisLambda,
         lwd=2,
         ty="l",
         xlab="t, time (days)",
         ylab="Heads per square metre per day",
         col="purple",
         main=sTmp,
         ylim=c(0,40),
     cex.lab=(11/12),
     cex.axis=(10/12),
     cex.main=(10/12))
  
    thisLambdaAerial <- rateChangeInfFromODESoln(odeRet,thisP,2)
    thisLambdaAphid <- rateChangeInfFromODESoln(odeRet,thisP,3)

    lines(odeRet$time,
        thisLambdaAerial,
        lwd=2,
        col="cyan4",
        lty=2)
    
    lines(odeRet$time,
        thisLambdaAphid,
        lwd=2,
        lty=2,
        col="pink")

    if(i == 1)
    {
    legend("topleft",
           c("Splash dispersed+Vectored","Splash dispersed","Vectored"),
           col=c("purple","cyan4","pink"),
           lwd=2,
           lty=c(1,2,2),cex=(9/12))
    }
}

plot(NULL, xlim=c(0,1), ylim=c(0,1), ylab="", xlab="",xaxt="n",yaxt="n",
     cex.lab=(11/12),
     cex.axis=(10/12),
     cex.main=(10/12),bty="n")

legend("top",
      c("Severity (NIV-like)", "AUDPC (NIV-like)", "Default value (NIV)", "Severity (DON-like)", "AUDPC (DON-like)", "Default value (DON)"), 
      col=c("darkgreen","lightgreen","black","darkblue","dodgerblue","grey"),
           lwd=2,
           ncol=1,
           lty=c(1,1,1,1,1),bty="n",cex=(9/12))    


allAlpha <- seq(0,1.0,by=0.05)
allV <- seq(0.0,2,by=0.025)
allR <- seq(0,0.5,by=0.01)
allDelta <- seq(0.0,2,by=0.025)
for(i in 1:4)
{
  if(i == 1)
  {
    allP <- allAlpha
    paramStr <- expression(alpha*", Acquisition probability")
    sLegStrNIV <- expression("Default "*alpha*" (NIV)")
    sLegStrDON <- expression("Default "*alpha*" (DON)")
    thisName <- "alpha"
    thisDefNIVVal <- defParamNIV$alpha
    thisDefDONVal <- defParamDON$alpha
    titleStr <- expression("Sensitivity to "*alpha)
  }
  if(i == 2)
  {
    allP <- allR
    paramStr <- expression(r*", Splash dispersal infection rate (/day)")
    sLegStrNIV <- expression("Default "*r*" (NIV)")
    sLegStrDON <- expression("Default "*r*" (DON)")
    thisName <- "r"
    thisDefNIVVal <- defParamNIV$r
    thisDefDONVal <- defParamDON$r
    titleStr <- expression("Sensitivity to "*r)
  }
  if(i == 3)
  {
    allP <- allV
#    paramStr <- expression(nu["-"]*", Unexposed aphid bias to infected heads")
    paramStr <- expression(nu*", Unexposed aphid bias to infected heads")
    sLegStrNIV <- expression("Default "*nu*" (NIV)")
    sLegStrDON <- expression("Default "*nu*" (DON)")
    thisName <- "nu"
    thisDefNIVVal <- defParamNIV$v
    thisDefDONVal <- defParamDON$v
    titleStr <- expression("Sensitivity to "*nu)
  }
  if(i == 4)
  {
    allP <- allDelta
#    paramStr <- expression(nu["-"]*", Unexposed aphid bias to infected heads")
    paramStr <- expression(delta*", Exposed aphid bias to healthy heads")
    sLegStrNIV <- expression("Default "*delta*" (NIV)")
    sLegStrDON <- expression("Default "*delta*" (DON)")
    thisName <- "delta"
    thisDefNIVVal <- defParamNIV$delta
    thisDefDONVal <- defParamDON$delta
    titleStr <- expression("Sensitivity to "*delta)
  }
  allNIVSev <- numeric(length(allP))
  allNIVAUDPC <- numeric(length(allP))
  allDONSev <- numeric(length(allP))
  allDONAUDPC <- numeric(length(allP))
  for(j in 1:length(allP))
  {
    thisParamNIV <- defParamNIV
    thisParamDON <- defParamDON
    
    if(thisName == "alpha")
    {
      thisParamNIV$alpha <- allP[j]
      thisParamDON$alpha <- allP[j]
    }
    
    if(thisName == "r")
    {
      thisParamNIV$r <- allP[j]
      thisParamDON$r <- allP[j]
    }
    
    if(thisName == "nu")
    {
      thisParamNIV$v <- allP[j]
      thisParamDON$v <- allP[j]
    }

    if(thisName == "delta")
    {
      thisParamNIV$delta <- allP[j]
      thisParamDON$delta <- allP[j]
    }
        
    odeRetNIV <- data.frame(ode(y = yZero, 
                           times = seq(0,tMax,length.out=101), 
                           func = fullModel, 
                           parms = thisParamNIV, 
                           method = "ode23"))
  
    thisLambdaNIV <- rateChangeInfFromODESoln(odeRetNIV,thisParamNIV,1)

    thisSevNIV <- severityFromLambda(odeRetNIV$time,thisLambdaNIV,thisParamNIV)

    thisAUDPCNIV <- integrate.xy(odeRetNIV$time, thisSevNIV)/(max(odeRetNIV$time)-min(odeRetNIV$time))
    
    odeRetDON <- data.frame(ode(y = yZero, 
                           times = seq(0,tMax,length.out=101), 
                           func = fullModel, 
                           parms = thisParamDON, 
                           method = "ode23"))

    thisLambdaDON <- rateChangeInfFromODESoln(odeRetDON,thisParamDON,1)

    thisSevDON <- severityFromLambda(odeRetDON$time,thisLambdaDON,thisParamDON)

    thisAUDPCDON <- integrate.xy(odeRetDON$time, thisSevDON)/(max(odeRetDON$time)-min(odeRetDON$time))
  
    allNIVSev[j] <- thisSevNIV[length(thisSevNIV)]
    allDONSev[j] <- thisSevDON[length(thisSevNIV)]
    allNIVAUDPC[j] <- thisAUDPCNIV
    allDONAUDPC[j] <- thisAUDPCDON

  }
  if(F)
  {
    print(thisName)
    print(allP)
    print(allNIVSev)
    print(allDONSev)
    print(allNIVAUDPC)
    print(allDONAUDPC)
  }  
  plot(allP,
       allNIVSev,
       col="darkgreen",
       xlab=paramStr,
       ylab="Severity (at 60 days) & AUDPC",
       main=titleStr,
       ty="l",
       lwd=2,
       ylim=c(0,1),
     cex.lab=(11/12),
     cex.axis=(10/12),
     cex.main=(11/12))
  
  lines(c(thisDefNIVVal,thisDefNIVVal),c(-1,2),lwd=2,col="black")
  lines(c(thisDefDONVal,thisDefDONVal),c(-1,2),lwd=2,col="grey")
 
  lines(allP,allNIVSev,lwd=2,col="darkgreen") # replot so overlays
  lines(allP,allDONSev,lwd=2,col="darkblue")
  lines(allP,allNIVAUDPC,lwd=2,col="lightgreen")
  lines(allP,allDONAUDPC,lwd=2,col="dodgerblue")
  
#  lines(c(thisDefNIVVal,thisDefNIVVal),c(-1,2),lwd=2,col="black")
#  lines(c(thisDefDONVal,thisDefDONVal),c(-1,2),lwd=2,col="grey")
}
```

Find and report severity and AUPDC for default values with NIV and DON

```{R echo=F}

thisParamNIV <- defParamNIV
thisParamDON <- defParamDON
odeRetNIV <- data.frame(ode(y = yZero, 
                       times = seq(0,tMax,length.out=101), 
                       func = fullModel, 
                       parms = thisParamNIV, 
                       method = "ode23"))

thisLambdaNIV <- rateChangeInfFromODESoln(odeRetNIV,thisParamNIV,1)

thisSevNIV <- severityFromLambda(odeRetNIV$time,thisLambdaNIV,thisParamNIV)

thisAUDPCNIV <- integrate.xy(odeRetNIV$time, thisSevNIV)/(max(odeRetNIV$time)-min(odeRetNIV$time))

odeRetDON <- data.frame(ode(y = yZero, 
                       times = seq(0,tMax,length.out=101), 
                       func = fullModel, 
                       parms = thisParamDON, 
                       method = "ode23"))

thisLambdaDON <- rateChangeInfFromODESoln(odeRetDON,thisParamDON,1)

thisSevDON <- severityFromLambda(odeRetDON$time,thisLambdaDON,thisParamDON)

thisAUDPCDON <- integrate.xy(odeRetDON$time, thisSevDON)/(max(odeRetDON$time)-min(odeRetDON$time))

print(paste("NIV: Sev =",thisSevNIV[length(thisSevNIV)],"AUDPC =",thisAUDPCNIV))
print(paste("DON: Sev =",thisSevDON[length(thisSevDON)],"AUDPC =", thisAUDPCDON))
```

## Parameterising within-spike severity

The following code generates the following figures (in order)

- Extended Data Figure 9. Panel (g)
- Extended Data Figure 9. Panel (h)

The results of this code, i.e., the best fitting parameters

- $\zeta_0 = 0.034$ (severity at time of first infection) 
- $\omega = 0.25$ day$^{-1}$ (rate of increase of severity). 

are used in generating the figures above.

Note that for two samples in the data file used for fitting the logistic response (Within Head Severity Data.xlsx) the severity decreases between 11 and 13 days. These samples are filtered out before doing the fit.

```{R echo=F,fitSevCurve,fig.height = 4,fig.align = 'center'}
logFunc <- function(time, zetaZero, omega)
{
  thisOmega <- 1 /(1 + (1/zetaZero - 1) * exp(-omega * time))
  return(thisOmega)
}

lsFitTwoParam <- function(v)
{
  paramOne <- v[1]
  paramTwo <- v[2]
  
  sumSq <- 0
  for(i in 1:length(allDataS))
  {
    thisE <- logFunc(allDataT[i],paramOne,paramTwo) - allDataS[i]
    sumSq <- sumSq + thisE * thisE  
  }
  return(sumSq)
}

newData <- read_excel("Within Head Severity Data.xlsx", sheet = "Sheet1")
par(las=1)
par(family=myFont)
par(mfrow = c(1,2))
plot(-100,
     -100,
     pch=20,
     col="red",
     main="Within-head severity",
     xlab=expression("Time post infection ("*tau*", days)"),
     ylab=expression("Normalised severity ( "*sigma(tau)*")"),
     ylim=c(0,1),
     xlim=c(0,30),
     cex.lab=(11/12),
     cex.axis=(10/12),
     cex.main=(10/12))

newData$`5dai Severity %` <- newData$`5dai Severity %`/100
newData$`7dai Severity %` <- newData$`7dai Severity %`/100
newData$`9dai Severity %` <- newData$`9dai Severity %`/100
newData$`11daiSeverity %` <- newData$`11daiSeverity %`/100
newData$`13dai Disease+Bleached Severity %` <- newData$`13dai Disease+Bleached Severity %`/100

goodIDX <- NULL
for(i in 1:nrow(newData))
{
  allX <- c(5,7,9,11,13)
  allY <- c(newData[i,]$`5dai Severity %`,newData[i,]$`7dai Severity %`,newData[i,]$`9dai Severity %`,newData[i,]$`11daiSeverity %`,newData[i,]$`13dai Disease+Bleached Severity %`)
  if(newData[i,]$`7dai Severity %` >= newData[i,]$`5dai Severity %`)
  {
    if(newData[i,]$`9dai Severity %` >= newData[i,]$`7dai Severity %`)
    {
      if(newData[i,]$`11daiSeverity %` >= newData[i,]$`9dai Severity %`)
      {
        if(newData[i,]$`13dai Disease+Bleached Severity %` >= newData[i,]$`11daiSeverity %`)
        {
          goodIDX <- c(goodIDX,i)
          lines(allX,allY,col="grey")
        }else{
          print(paste("Filtered sample",newData[i,]$Sample,"since severity drops from 11 to 13 days"))
        }
      }else{
          print(paste("Filtered sample",newData[i,]$Sample,"since severity drops from 9 to 11 days"))
      }
    }else{
      print(paste("Filtered sample",newData[i,]$Sample,"since severity drops from 7 to 9 days"))
    }
  }else{
    print(paste("Filtered sample",newData[i,]$Sample,"since severity drops from 5 to 7 days"))
  }
}

myStdErr <- function(x)
{
  return(sqrt(var(x)/length(x)))
}

allDataS <- c(mean(newData$`5dai Severity %`[goodIDX]),
             mean(newData$`7dai Severity %`[goodIDX]),
             mean(newData$`9dai Severity %`[goodIDX]),
             mean(newData$`11daiSeverity %`[goodIDX]),
             mean(newData$`13dai Disease+Bleached Severity %`[goodIDX]))
allDataStdE <- c(myStdErr(newData$`5dai Severity %`[goodIDX]),
                 myStdErr(newData$`7dai Severity %`[goodIDX]),
                 myStdErr(newData$`9dai Severity %`[goodIDX]),
                 myStdErr(newData$`11daiSeverity %`[goodIDX]),
                 myStdErr(newData$`13dai Disease+Bleached Severity %`[goodIDX]))

allDataT <- seq(5,13,by=2)
for(i in 1:length(allDataS))
{
  points(allDataT[i],allDataS[i],cex=2,col="black",pch=20)
  lines(c(allDataT[i],allDataT[i]),c(allDataS[i]-1.96*allDataStdE[i],allDataS[i]+1.96*allDataStdE[i]),lwd=3)
  lines(c(allDataT[i]-0.25,allDataT[i]+0.25),c(allDataS[i]-1.96*allDataStdE[i],allDataS[i]-1.96*allDataStdE[i]),lwd=3)
  lines(c(allDataT[i]-0.25,allDataT[i]+0.25),c(allDataS[i]+1.96*allDataStdE[i],allDataS[i]+1.96*allDataStdE[i]),lwd=3)
}
t <- seq(0,30,by=1)

vTmp <- optim(c(0.1,0.2),lsFitTwoParam)
vTmpTwo <- vTmp

lines(t,
     logFunc(t,vTmp$par[1],vTmp$par[2]),
     lwd=3,
     col="dodgerblue")

legend("bottomright",
       legend=c("Data","Fit",expression("Mean"%+-%"SE")),
       col=c("grey","dodgerblue","black"),
       lwd=c(1,3,3),
       pch=c(NA,NA,20),cex=(9/12))

allZeta <- seq(0.01,0.1,by=0.01/2)
nZeta <- length(allZeta)
allOmega <- seq(0.05,0.4,by=0.05/2)
nOmega <- length(allOmega)
allLS <- matrix(NA,nrow=nZeta,ncol=nOmega)
for(i in 1:nZeta)
{
  for(j in 1:nOmega)
  {
    v <- c(allZeta[i],allOmega[j])
    allLS[i,j] <- lsFitTwoParam(v)
  }
}

rf <- colorRampPalette(rev(brewer.pal(9,'YlGnBu')))   # make colors
r <- rf(64)

image.plot(allZeta,
           allOmega,
           allLS,
           xlab=expression("Initial severity ("*zeta[0]*")"),
           ylab=expression("Rate of increase ("*omega*")"),
           main="Sum of squares",
           cex.lab=(11/12),
           cex.axis=(10/12),
           cex.main=(10/12),
           col=r,
           legend.lab="SS",
           axis.args=list(cex.axis=0.7))
points(vTmp$par[1],vTmp$par[2],pch=20,col="white",cex=2)
print(vTmp)

print("Rounded to 2 significant figures, the best fitting parameters are:")
s<-sprintf("Zeta0=%.3f",vTmp$par[1])
print(s)
s<-sprintf("Omega=%.2f",vTmp$par[2])
print(s)

```

